Improved state-space models for inference about spatial and temporal variation in abundance from count data

### 2014-08-04 improved_state_space.tex
Removed comments

\textit{Key words}: abundance, Dail and Madsen model, density-dependence,
detectability, Gompertz model, immigration, %-logistic
population dynamics, Ricker model, zero-inflated %-logistic

%\vspace{0.5cm}

\section*{Introduction}
Models of population dynamics are vital in both theoretical and
applied ecological research. For instance, 
the importance and existence of phenomenon such as
density-dependent population regulation, population cycling, and
spatial synchrony have been studied by comparing data from natural
populations with theoretical models
\citep{may:1975,turchin:1990,bjornstad_etal:1999}. 
In applied contexts, population models are used for estimating extinction risk 
\citep{nadeem_lele:2011,hostetler_etal:2012} and for predicting the %schoener_spiller:1992,
effects of future environmental conditions or conservation actions on
population size \citep{jamieson_brooks:2004,hatfield_etal:2012}.

Two challenges are routinely encountered when trying to fit models %Several
of population dynamics to data from field studies. 
First, deterministic models of population dynamics are generally inadequate due to process %virtually always
variation, the stochasticity in demographic parameters and environmental % inherent
conditions \citep{bjornstad_grenfell:2001,saether_engen:2002}.
Second, abundance---the natural state variable in studies
of population dynamics---can rarely be observed perfectly in field
studies because of observation error, such as imperfect
detection \citep{link_nichols:1994,kery_etal:2009}.

@@

A simple state-space model can be described as follows.
Let $N_t$ be the abundance of a species during year $t$, for
$t=1,\hdots,T$, and let $X_t$ be
the observed data, which differs from $N_t$ due to observation error,
a random effect denoted $\zeta_t$. Temporal variation in $N_t$ is
modeled using a population growth model, $\mu(N_{t-1})$
and random process variation denoted $\eta_t$.
The growth model may be density-dependent, as in the case of the 
Ricker-logistic model (henceforth Ricker model), or it might be 
density-independent, such as when growth
is exponential. %% Exponential or geometric?
A generic model can be written: 
\begin{linenomath*}
\begin{gather}
  \label{eq:ss1}
  \begin{align}
    N_1 &= x_0 \nonumber \\ %X_1
N_t &= \mu(N_{t-1}) + \eta_{t-1} \quad \text{for} \; 
t=2,\hdots,T  \\
X_t &= N_t + \zeta_t \qquad \qquad \;\, \text{for} \;
t=1,\hdots,T. \nonumber 
  \end{align}
\end{gather}
\end{linenomath*}
where $x_0$ is the distribution for initial abundance.  In classical
state-space models, the two sets of random effects
are assumed to be i.i.d. Gaussian variables: 
$\eta_t \sim \mathrm{N}(0, \sigma)$ and
$\zeta_t \sim \mathrm{N}(0, \tau)$. 
The process variation associated with $N_1$ is often ignored ($x_0 = X_1$), or
the population is assumed to be at stochastic equilibrium such that the $x_0$ can
be regarded as an outcome of the equilibrium distribution \citep{devalpine_hastings:2002}.

@@

Several extensions of state-space models have been proposed to
overcome the limitations described above. \citet{devalpine_hastings:2002} and
\citet{brooks_etal:2004} described methods for fitting models with non-Gaussian
distributions for the process and observation errors. An observation model %with
%more intuitive interpretations, such as those that 
explicitly modeling
detection probability has been proposed by 
\citet{kery_etal:2009}. \citet{lele_etal:1998} and 
\citet{kery_etal:2009} developed models allowing for inference about
spatial and temporal variation in abundance, and their developments
also resolved the problems of non-identifiability for the parameters
of the initial state at time $t=1$. Of these extensions, only
the work by \citet{kery_etal:2009} addressed several of these limitations
simultaneously; however, their model did not include serial
dependence, which is a hallmark of classical population growth models as well as state-space models.

@@

The DM model assumes that abundance at time $t>1$ is a function of
abundance at time $t-1$, i.e. abundance at each site evolves as a
first order Markov process. 
\citet{dail_madsen:2011} considered several models to describe the temporal dynamics;
however, in each case they modeled $N_t$ as the sum of two random variables:
$S_{i,t}$, the number of individuals surviving from $t-1$ and not
emigrating; and $G_{i,t}$ the number of new individuals entering
the population. Their most general model was
\begin{linenomath*}
\begin{equation}
\left.\begin{aligned}
S_{i,t}|N_{i,t-1} &\sim \mathrm{Bin}(N_{i,t-1}, \omega) \\
G_{i,t}|N_{i,t-1} &\sim \mathrm{Pois}(\gamma(N_{i,t-1})) \\
N_{i,t} &= S_{i,t}+G_{i,t}
\end{aligned}\right\} \quad \text{for} \; t=2,\hdots,T
\label{eq:Nt}
\end{equation}
\end{linenomath*}
where $\omega$ is the apparent survival probability and $\gamma$
is the recruitment rate, which can depend on $N_{i,t-1}$.
\citet{dail_madsen:2011} proposed three
models for $\gamma$: the constant model,
$G_{i,t} \sim \mathrm{Pois}(\gamma)$ where recruitment does not
depend on $N_{i,t-1}$, and which simulates a ``propagule rain'' of new %
individuals; the autoregressive model, $G_{i,t} \sim
\mathrm{Pois}(\gamma N_{i,t-1})$
of geometric or density independent growth; and the
no-trend model, $\gamma = (1-\omega)\Lambda$, which assumes the
population is at equilibrium. Covariates of
$\omega$ and $\gamma$ can be accommodated, e.g., % easily %for example
using logit- and log-linear models respectively.

@@

\section*{Model Extensions}
\label{sec:ext}
\subsection*{Population growth models}
%\begin{samepage}
Partitioning population growth into survival and recruitment
is useful  
in terms of providing a mechanistic description of population
dynamics, but this is not always possible using simple count data,
especially when the sites are not closed with respect to immigration
and emigration. 
When the mechanistic model is unrealistic, we suggest replacing it
with classical population growth models that have a long history in ecology.
The exponential growth model is the simplest: %% Should this geometric instead of exponential?
$N_{i,t} = N_{i,t-1}e^r$ where $r$ is the intrinsic
rate of increase. 
To allow for demographic stochasticity, we can regard $N_{i,t}$ as a Poisson 
random variable, simplifying Eq.~\ref{eq:Nt} to:
\begin{linenomath*}
\begin{equation}
  N_{i,t} \sim \mathrm{Pois}(\exp(r)N_{i,t-1}).
\label{eq:exp}
\end{equation}
\end{linenomath*}

Density-dependent versions of the model are also possible.  For
example:
\begin{linenomath*}
\begin{equation}
  N_{i,t} \sim \mathrm{Pois}(N_{i,t-1}\exp(r(1-N_{i,t-1}/K)))
\label{eq:rick}
\end{equation}
\end{linenomath*}
where $K$ is the stable equilibrium of the population and $r$ is
the instantaneous population growth rate at low population
densities, and both parameters are constrained to be positive. This is
a stochastic version of the 
Ricker model \citep{ricker:1954}. Another option is a 
modified Gompertz-logistic
density-dependent model \citep[henceforth Gompertz model,][]{hart_gotelli:2011}:
\begin{linenomath*}
\begin{equation}
N_{i,t} \sim \mathrm{Pois}(N_{i,t-1}\exp(r(1-\log(N_{i,t-1}+1)/\log(K+1))))
\label{eq:gomp}
\end{equation}
\end{linenomath*}
Here the interpretations of $r$ and $K$ are similar to those in the
Ricker model. Because a single Poisson distribution controls
the distribution of $N_{i,t}$ in each of these models, the discrete 
convolution used by \citet{dail_madsen:2011} to construct the
likelihood is not required, speeding up processing time.
%\end{samepage}

@@

\subsection*{Environmental and demographic stochasticity}
\label{subsec:stoch}
The original DM models allows for demographic stochasticity --- variation in population growth 
rate due to the randomness of birth and death processes --- 
via the binomial and Poisson distributions for apparent survival and recruitment.  Our population
growth extensions use a single Poisson distribution for abundance. 
% NOTE: I would save this for the discussion
Conceptually, this is an appropriate method of modeling demographic stochasticity 
 for semelparous organisms with one generation per sampling interval (year)
 \citep{bonsall_hastings:2004},
and the DM models appropriate for modeling the demographic stochasticity of iteroparous organisms,
if they are underdispersed compared to the Poisson (but see Discussion).
% but may actually perform better for iteroparous organisms as well (see Discussion).  
 Dynamics of semelparous organisms with more than one generation per year 
 (many insects) may be overdispersed compared to the Poisson.  If the number of generations per
 year is known, then this can be accounted for in our models by adding extra $N$'s between each
 count (see Supplement 1).

One way to model environmental stochasticity --- variation in population growth
rate due to stochastic environmental conditions --- is by explicitly
modeling dynamic parameters as functions of environmental covariates. %, such as weather.  
For example, in the Gompertz model (Eq.~\ref{eq:gomp}), $r$ could be a 
function of temperature  
and $K$ a function of rainfall.

@@

Current BBS trend estimators deal with this problem by
treating observer identity as a random 
effect \citep{link_sauer:2002,sauer_link:2011}.
To include random observer effects in DM models, 
Eq.~\ref{eq:p1} can be modified to:
\begin{linenomath*}
\begin{gather}
X_{i,j,t} \sim \mathrm{Bin}(N_{i,t}, p_j) \nonumber \\
\mathrm{logit}(p_j) \sim \mathrm{N}(\mu_p, \sigma_p)
\label{eq:pobs}
\end{gather}
\end{linenomath*}
where $X_{i,j,t}$ is the number of individuals counted at site $i$ by
observer $j$ in year $t$, $p_j$ is observer-specific detection probability,
$\mu_p$ is the mean detection probability (on the logit scale), and $\sigma_p$ is
the logit-scale standard deviation of the random observer effect. % (also on the logit scale).

@@

%In practice, 
Simulation methods such as Markov chain Monte Carlo (MCMC)
are used to estimate the posterior distributions. While time %simulate
consuming, MCMC may be the only viable approach for estimating
parameters, especially for hierarchical models with many random
effects.  
Software packages such as 
\textbf{JAGS} \citep{plummer:2003} make these methods readily
available to ecologists, and we have provided examples in 
Supplement 1.

@@

We applied our model extensions to North American Breeding Bird Survey
(BBS) data collected from 1966-2010 in 
Maryland and Virginia. For our focal species, we selected the
ovenbird (\textit{Seiurus aurocapilla}), an abundant, %and widespread,
forest-breeding migratory songbird with a stable or increasing trend
in the region \citep{porneluzi_etal:2011}.

@@

The negative binomial distribution was strongly supported for ovenbird
initial abundance over the Poisson and zero-inflated Poisson
(Table B1A in Appendix B). 
The best supported model for $p$ included effects of wind speed %additive 
and an observer's first run of a route (Table B1B, model B.1). First run 
and increasing wind speeds both decreased $p$. All dynamics models 
with immigration were better supported than models without immigration 
(Table B1C); the best supported was Ricker + immigration % of these the
(see Appendix B, Fig. B1 for estimates).
Constant-recruitment dynamics models estimated unrealistically high
survival probabilities ($\omega$ = 1 $\pm$ 1.3e-05), whereas 
the geometric- recruitment and geometric-recruitment + immigration
models estimated nonviably low survival probabilities 
($\omega$ =
0.058 $\pm$ 0.046 and 0.026 $\pm$ 0.060, respectively).

@@

One strategy for adhering to the original formulation of the
model that included survival and recruitment parameters would be to
use either prior information or additional data \citep{zipkin_etal:2014}. For example, in a 
Bayesian analysis existing estimates of demographic parameters
could be used as informative priors, which might make it possible to
seperate the effects of movement versus births and deaths.
Alternatively, ecologists could combine count data with demographic data,
%from, for example,
%capture-recapture data. Models that are fitted to count data and
%demographic data are 
often referred to as an integrated population model
\citep[IPM;][]{besbeas_etal:2002,schaub_etal:2007,chandler_clark:2014}.
We believe that a DM modeling framework could work well in this context,
and that the DM model could be viewed as a specific case of the
IPM proposed by \citet{buckland_etal:2004}.
However, we caution that when count data are affected by
movement and vital rates, it would be unwise to ignore the movement
processes.

>>>>

\textit{Key words}: abundance, Dail and Madsen model, density-dependence,
detectability, Gompertz model, immigration, 
population dynamics, Ricker model, zero-inflated

\section*{Introduction}
Models of population dynamics are vital in both theoretical and
applied ecological research. For instance, 
the importance and existence of phenomenon such as
density-dependent population regulation, population cycling, and
spatial synchrony have been studied by comparing data from natural
populations with theoretical models
\citep{may:1975,turchin:1990,bjornstad_etal:1999}. 
In applied contexts, population models are used for estimating extinction risk 
\citep{nadeem_lele:2011,hostetler_etal:2012} and for predicting the 
effects of future environmental conditions or conservation actions on
population size \citep{jamieson_brooks:2004,hatfield_etal:2012}.

Two challenges are routinely encountered when trying to fit models 
of population dynamics to data from field studies. 
First, deterministic models of population dynamics are generally inadequate due to process 
variation, the stochasticity in demographic parameters and environmental 
conditions \citep{bjornstad_grenfell:2001,saether_engen:2002}.
Second, abundance---the natural state variable in studies
of population dynamics---can rarely be observed perfectly in field
studies because of observation error, such as imperfect
detection \citep{link_nichols:1994,kery_etal:2009}.

@@

A simple state-space model can be described as follows.
Let $N_t$ be the abundance of a species during year $t$, for
$t=1,\hdots,T$, and let $X_t$ be
the observed data, which differs from $N_t$ due to observation error,
a random effect denoted $\zeta_t$. Temporal variation in $N_t$ is
modeled using a population growth model, $\mu(N_{t-1})$
and random process variation denoted $\eta_t$.
The growth model may be density-dependent, as in the case of the 
Ricker-logistic model (henceforth Ricker model), or it might be 
density-independent, such as when growth
is exponential. 
A generic model can be written: 
\begin{linenomath*}
\begin{gather}
  \label{eq:ss1}
  \begin{align}
    N_1 &= x_0 \nonumber \\ 
N_t &= \mu(N_{t-1}) + \eta_{t-1} \quad \text{for} \; 
t=2,\hdots,T  \\
X_t &= N_t + \zeta_t \qquad \qquad \;\, \text{for} \;
t=1,\hdots,T. \nonumber 
  \end{align}
\end{gather}
\end{linenomath*}
where $x_0$ is the distribution for initial abundance.  In classical
state-space models, the two sets of random effects
are assumed to be i.i.d. Gaussian variables: 
$\eta_t \sim \mathrm{N}(0, \sigma)$ and
$\zeta_t \sim \mathrm{N}(0, \tau)$. 
The process variation associated with $N_1$ is often ignored ($x_0 = X_1$), or
the population is assumed to be at stochastic equilibrium such that the $x_0$ can
be regarded as an outcome of the equilibrium distribution \citep{devalpine_hastings:2002}.

@@

Several extensions of state-space models have been proposed to
overcome the limitations described above. \citet{devalpine_hastings:2002} and
\citet{brooks_etal:2004} described methods for fitting models with non-Gaussian
distributions for the process and observation errors. An observation model 
explicitly modeling
detection probability has been proposed by 
\citet{kery_etal:2009}. \citet{lele_etal:1998} and 
\citet{kery_etal:2009} developed models allowing for inference about
spatial and temporal variation in abundance, and their developments
also resolved the problems of non-identifiability for the parameters
of the initial state at time $t=1$. Of these extensions, only
the work by \citet{kery_etal:2009} addressed several of these limitations
simultaneously; however, their model did not include serial
dependence, which is a hallmark of classical population growth models as well as state-space models.

@@

The DM model assumes that abundance at time $t>1$ is a function of
abundance at time $t-1$, i.e. abundance at each site evolves as a
first order Markov process. 
\citet{dail_madsen:2011} considered several models to describe the temporal dynamics;
however, in each case they modeled $N_t$ as the sum of two random variables:
$S_{i,t}$, the number of individuals surviving from $t-1$ and not
emigrating; and $G_{i,t}$ the number of new individuals entering
the population. Their most general model was
\begin{linenomath*}
\begin{equation}
\left.\begin{aligned}
S_{i,t}|N_{i,t-1} &\sim \mathrm{Bin}(N_{i,t-1}, \omega) \\
G_{i,t}|N_{i,t-1} &\sim \mathrm{Pois}(\gamma(N_{i,t-1})) \\
N_{i,t} &= S_{i,t}+G_{i,t}
\end{aligned}\right\} \quad \text{for} \; t=2,\hdots,T
\label{eq:Nt}
\end{equation}
\end{linenomath*}
where $\omega$ is the apparent survival probability and $\gamma$
is the recruitment rate, which can depend on $N_{i,t-1}$.
\citet{dail_madsen:2011} proposed three
models for $\gamma$: the constant model,
$G_{i,t} \sim \mathrm{Pois}(\gamma)$ where recruitment does not
depend on $N_{i,t-1}$, and which simulates a ``propagule rain'' of new 
individuals; the autoregressive model, $G_{i,t} \sim
\mathrm{Pois}(\gamma N_{i,t-1})$
of geometric or density independent growth; and the
no-trend model, $\gamma = (1-\omega)\Lambda$, which assumes the
population is at equilibrium. Covariates of
$\omega$ and $\gamma$ can be accommodated, e.g., 
using logit- and log-linear models respectively.

@@

\section*{Model Extensions}
\label{sec:ext}
\subsection*{Population growth models}
Partitioning population growth into survival and recruitment
is useful  
in terms of providing a mechanistic description of population
dynamics, but this is not always possible using simple count data,
especially when the sites are not closed with respect to immigration
and emigration. 
When the mechanistic model is unrealistic, we suggest replacing it
with classical population growth models that have a long history in ecology.
The exponential growth model is the simplest: 
$N_{i,t} = N_{i,t-1}e^r$ where $r$ is the intrinsic
rate of increase. 
To allow for demographic stochasticity, we can regard $N_{i,t}$ as a Poisson 
random variable, simplifying Eq.~\ref{eq:Nt} to:
\begin{linenomath*}
\begin{equation}
  N_{i,t} \sim \mathrm{Pois}(\exp(r)N_{i,t-1}).
\label{eq:exp}
\end{equation}
\end{linenomath*}

Density-dependent versions of the model are also possible.  For
example:
\begin{linenomath*}
\begin{equation}
  N_{i,t} \sim \mathrm{Pois}(N_{i,t-1}\exp(r(1-N_{i,t-1}/K)))
\label{eq:rick}
\end{equation}
\end{linenomath*}
where $K$ is the stable equilibrium of the population and $r$ is
the instantaneous population growth rate at low population
densities, and both parameters are constrained to be positive. This is
a stochastic version of the 
Ricker model \citep{ricker:1954}. Another option is a 
modified Gompertz-logistic
density-dependent model \citep[henceforth Gompertz model,][]{hart_gotelli:2011}:
\begin{linenomath*}
\begin{equation}
N_{i,t} \sim \mathrm{Pois}(N_{i,t-1}\exp(r(1-\log(N_{i,t-1}+1)/\log(K+1))))
\label{eq:gomp}
\end{equation}
\end{linenomath*}
Here the interpretations of $r$ and $K$ are similar to those in the
Ricker model. Because a single Poisson distribution controls
the distribution of $N_{i,t}$ in each of these models, the discrete 
convolution used by \citet{dail_madsen:2011} to construct the
likelihood is not required, speeding up processing time.

@@

\subsection*{Environmental and demographic stochasticity}
\label{subsec:stoch}
The original DM models allows for demographic stochasticity --- variation in population growth 
rate due to the randomness of birth and death processes --- 
via the binomial and Poisson distributions for apparent survival and recruitment.  Our population
growth extensions use a single Poisson distribution for abundance. 
Conceptually, this is an appropriate method of modeling demographic stochasticity 
 for semelparous organisms with one generation per sampling interval (year)
 \citep{bonsall_hastings:2004},
and the DM models appropriate for modeling the demographic stochasticity of iteroparous organisms,
if they are underdispersed compared to the Poisson (but see Discussion).
 Dynamics of semelparous organisms with more than one generation per year 
 (many insects) may be overdispersed compared to the Poisson.  If the number of generations per
 year is known, then this can be accounted for in our models by adding extra $N$'s between each
 count (see Supplement 1).

One way to model environmental stochasticity --- variation in population growth
rate due to stochastic environmental conditions --- is by explicitly
modeling dynamic parameters as functions of environmental covariates. 
For example, in the Gompertz model (Eq.~\ref{eq:gomp}), $r$ could be a 
function of temperature  
and $K$ a function of rainfall.

@@

Current BBS trend estimators deal with this problem by
treating observer identity as a random 
effect \citep{link_sauer:2002,sauer_link:2011}.
To include random observer effects in DM models, 
Eq.~\ref{eq:p1} can be modified to:
\begin{linenomath*}
\begin{gather}
X_{i,j,t} \sim \mathrm{Bin}(N_{i,t}, p_j) \nonumber \\
\mathrm{logit}(p_j) \sim \mathrm{N}(\mu_p, \sigma_p)
\label{eq:pobs}
\end{gather}
\end{linenomath*}
where $X_{i,j,t}$ is the number of individuals counted at site $i$ by
observer $j$ in year $t$, $p_j$ is observer-specific detection probability,
$\mu_p$ is the mean detection probability (on the logit scale), and $\sigma_p$ is
the logit-scale standard deviation of the random observer effect.

@@

Simulation methods such as Markov chain Monte Carlo (MCMC)
are used to estimate the posterior distributions. While time 
consuming, MCMC may be the only viable approach for estimating
parameters, especially for hierarchical models with many random
effects.  
Software packages such as 
\textbf{JAGS} \citep{plummer:2003} make these methods readily
available to ecologists, and we have provided examples in 
Supplement 1.

@@

We applied our model extensions to North American Breeding Bird Survey
(BBS) data collected from 1966-2010 in 
Maryland and Virginia. For our focal species, we selected the
ovenbird (\textit{Seiurus aurocapilla}), an abundant, 
forest-breeding migratory songbird with a stable or increasing trend
in the region \citep{porneluzi_etal:2011}.

@@

The negative binomial distribution was strongly supported for ovenbird
initial abundance over the Poisson and zero-inflated Poisson
(Table B1A in Appendix B). 
The best supported model for $p$ included effects of wind speed  
and an observer's first run of a route (Table B1B, model B.1). First run 
and increasing wind speeds both decreased $p$. All dynamics models 
with immigration were better supported than models without immigration 
(Table B1C); the best supported was Ricker + immigration 
(see Appendix B, Fig. B1 for estimates).
Constant-recruitment dynamics models estimated unrealistically high
survival probabilities ($\omega$ = 1 $\pm$ 1.3e-05), whereas 
the geometric- recruitment and geometric-recruitment + immigration
models estimated nonviably low survival probabilities 
($\omega$ =
0.058 $\pm$ 0.046 and 0.026 $\pm$ 0.060, respectively).

@@

One strategy for adhering to the original formulation of the
model that included survival and recruitment parameters would be to
use either prior information or additional data \citep{zipkin_etal:2014}. For example, in a 
Bayesian analysis existing estimates of demographic parameters
could be used as informative priors, which might make it possible to
seperate the effects of movement versus births and deaths.
Alternatively, ecologists could combine count data with demographic data,
often referred to as an integrated population model
\citep[IPM;][]{besbeas_etal:2002,schaub_etal:2007,chandler_clark:2014}.
We believe that a DM modeling framework could work well in this context,
and that the DM model could be viewed as a specific case of the
IPM proposed by \citet{buckland_etal:2004}.
However, we caution that when count data are affected by
movement and vital rates, it would be unwise to ignore the movement
processes.
